---
title: "Reproducible Research Week 2 Course Project 1"
author: "Author: Tumisang Liphoto"
date: "Date: 12/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R markdown file that presents the solutions to the questions asked in
the Reproducible Research Week 2 Course Project 1. The data used used in this solution
is downloaded from [this][1] address on 12/13/2021.

The data is in a form of a zip file that gets decompressed.The task here comprises of 5 steps, namely:

#. Loading and pre-processing the data.
#. Histogram of the total number of steps taken each day
#. Calculate and report the mean and median of the total number of steps taken per day.
#. Create a time series plot of the data.
#. Finding 5-minute interval that, on average, contains the maximum number of steps
#. Imputting missing values
    i. Strategy for filling in all of the missing values.
    ii. Create new dataset without missing values.
    iii. Make a histogram of the total number of steps taken each day and Calculate and report the mean and            median total number of steps taken per day
     a) Do these values differ from the estimates from the first part of the assignment?
     b) What is the impact of imputing missing data on the estimates of the total daily           number of steps?

# 1. Loading and Preprocessing Data

```{r getData}
library(data.table) 
library(dplyr)#load dplyr
library(ggplot2)#load ggplot2
library(magrittr)

dataFile <- "Reproducible_Research _Week2_Peer_Review"

if (!file.exists(dataFile)){
  fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  print("downloading....")
  download.file(fileURL, destfile = dataFile, method="curl")
  print("done download")
  print("unzipping")
  unzip (zipfile = dataFile)
  print("done unzipping")
}

activitydata <- read.csv("activity.csv", header = TRUE)
head(activitydata)
```


```{r summary}
  summary(activitydata)
```
# 2. Histogram of the total number of steps taken each day
```{r stepsEachDay}
dailyData <- activitydata %>% select(date, steps) %>% group_by(date) %>% summarize(totalSteps= sum(steps)) %>%na.omit()

hist(dailyData$totalSteps, xlab = "Daily Steps",main=" Total steps by day.", breaks = 20, col = "red")
```

# 3. Calculate and report the mean and median of the total number of steps taken per day.
```{r mean}
mean(dailyData$totalSteps)
```
 

```{r median}
median(dailyData$totalSteps)
```

# 4. Create a time series plot of the data.
```{r timeseries}
timeSeries <- activitydata %>% select(interval, steps) %>%na.omit() %>% group_by(interval) %>% summarize(totalSteps= mean(steps))  
ggplot( timeSeries, aes(x=interval, y=totalSteps), col("red"))+ geom_line(colour = "red") + ggtitle("Taken Steps Time Series.") + xlab("Intervals") + ylab("Total Steps")+ theme(plot.title = element_text(hjust = 0.5))
```
# 5. Finding 5-minute interval that, on average, contains the maximum number of steps
```{r 5 minutes}
  timeSeries[which(timeSeries$totalSteps== max(timeSeries$totalSteps)),]
```
# 6. Calculate and report the total number of missing values.
```{r missing values }
 #sum(is.na(activitydata))

 impune <- function(Data) replace(Data, is.na(Data), mean(Data, na.rm = TRUE))
 
 averagedData <- activitydata %>% group_by(interval) %>% mutate(steps = impune(steps))
 
 head(averagedData)
 #sum(is.na(averagedData))
```

```{r summary2}
  summary(averagedData)
```

# 7. Histogram of the total number of steps taken each day - Missing data averaged (include mean and median)
```{r removed missing values - histogram}

  dailyData <- averagedData %>% select(date, steps) %>% group_by(date) %>% summarize(totalSteps= sum(steps))
  hist(dailyData$totalSteps, xlab = "Daily Steps",main=" Total steps by day.", breaks = 20, col = "blue")
```

# 3. Calculate and report the mean and median of the total number of steps taken per day.
```{r mean and median with nas removed}
mean(dailyData$totalSteps)
median(dailyData$totalSteps)
```
## Are there differences in activity patterns between weekdays and weekends?
The missing values have no impact on the calculations of both the mean and median as these are excluded when
calculating the two values.

```{r}
averagedData$date <- as.Date(averagedData$date)
averagedData$weekday <- weekdays(averagedData$date)
averagedData$weekend <- ifelse(averagedData$weekday=="Saturday" | averagedData$weekday=="Sunday", "Weekend", "Weekday" )
```

```{r}
meandataweekendweekday <- aggregate(averagedData$steps , by= list(averagedData$weekend, averagedData$interval), na.omit(mean))
names(meandataweekendweekday) <- c("weekend", "interval", "steps")

ggplot(meandataweekendweekday, aes(x=interval, y=steps, color=weekend)) + geom_line()+
facet_grid(weekend ~.) + xlab("Interval") + ylab("Mean of Steps") +
    ggtitle("Comparison of Average Number of Steps in Each Interval")
```



[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip  "this"
